A sample workflow

### 2025-04-25 analyses/african_mammals_workflow.qmd
retro and inference on simulated data with marginal means

### Simulating data
Now we start to simulate!
*Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.* 
```{r sim data}

@@

@@

dataplot <- data.frame(
  y, 
  year = years - 1980,
  species = as.character(spid_perobs),
  population = paste0(spid_perobs,popid_perobs)
)

dataplot$logy <- log(dataplot$y)

ggplot(data = dataplot) +
  geom_line(aes(x = year+1980, y = y, 
                group = paste0(species, population),
                color = species)) +
  theme_bw()

ggplot(data = dataplot) +
  geom_line(aes(x = year+1980, y = logy, 
                group = paste0(species, population),
                color = species)) +
  theme_bw()

@@

nestlm <- lm(logy ~ species/population + year * species/population, data = dataplot)
coefs <- summary(nestlm)$coefficients

@@

retrodata <- data.frame(obs = dataplot$logy, pred = predict(nestlm),
                        year = dataplot$year,
                        species = dataplot$species, id = dataplot$population)

ggplot(data = retrodata) +
  facet_wrap(~ paste0("Species ", species), scales = "free") +
  geom_point(aes(x = year, y = obs, group = id, color = as.character(id)),
             size = 1) + 
  geom_line(aes(x = year, y = pred, group = id, color = as.character(id))) +
  theme_bw() + theme(legend.position = 'none')

ggplot(data = retrodata) +
  facet_wrap(~ paste0("Species ", species), scales = "free") +
  geom_point(aes(x = year, y = exp(obs), group = id, color = as.character(id)),
             size = 1) + 
  geom_line(aes(x = year, y = exp(pred), group = id, color = as.character(id))) +
  theme_bw() + theme(legend.position = 'none')

@@

The model does a poor job at recovering species-level parameters.

```{r inference sim data}
species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("species", sp), "Std. Error"]}))

@@

slopes <- data.frame(obs=mu_species_beta, est=species_slope, est.sd=species_slope_sd, species = as.character(1:nspecies))
ggplot(data = slopes) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
  geom_pointrange(aes(ymin = est-est.sd, ymax = est+est.sd, x = obs, y = est, color = species)) +
  theme_bw()
```

@@

@@

```{r}
#| eval: false

@@

lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,]
lpi_subset_rsh$year <- lpi_subset_rsh$year-1980
lpi_subset_rsh$logvalue <- log(lpi_subset_rsh$value)
lpi_subset_rsh$ID <- as.character(lpi_subset_rsh$ID)

>>>>

### Simulating data
Now we start to simulate!
*Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.*

```{r sim data}

@@

# save simulated intercepts and slopes
pops_fit <- data.frame()
pid <- 1
for(sp in 1:nspecies){
  for(p in 1:npops_persp[sp]){
    pops_fit <- rbind(pops_fit, data.frame(species = sp, pop = pid))
    pid <- pid + 1
  }
}
pops_fit$int <- mu_pop_alpha
pops_fit$slp <- mu_pop_beta

@@

datasim <- data.frame(
  y, 
  year = years - 1980,
  species = as.character(spid_perobs),
  population = paste0(spid_perobs,popid_perobs)
)

datasim$logy <- log(datasim$y)

ggplot(data = datasim) +
  geom_line(aes(x = year+1980, y = y, 
                group = population,
                color = species)) +
  theme_bw() +
  labs(x = 'Year', y = 'Simulated counts', colour = 'Species')

ggplot(data = datasim) +
  geom_line(aes(x = year+1980, y = logy, 
                group = population,
                color = species)) +
  theme_bw() +
  labs(x = 'Year', y = 'Simulated counts (log-scale)', colour = 'Species')

@@

nestlm <- lm(logy ~ species/population + year * species/population, data = datasim)
coefs <- summary(nestlm)$coefficients

@@

retrodata <- data.frame(obs = datasim$logy, pred = predict(nestlm),
                        year = datasim$year,
                        species = datasim$species, id = datasim$population)

ggplot(data = retrodata) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  facet_wrap(~ paste0("Species ", species), scales = "free") +
  geom_point(aes(x = obs, y = pred, group = id, color = as.character(id)),
             size = 0.5) + 
  theme_bw() + theme(legend.position = 'none') +
  labs(x = 'Simulated counts (log-scale)', y = 'Estimated counts (log-scale)')

ggplot(data = retrodata) +
  facet_wrap(~ paste0("Species ", species), scales = "free") +
  geom_point(aes(x = year, y = obs, group = id, color = as.character(id)),
             size = 1) + 
  geom_line(aes(x = year, y = pred, group = id, color = as.character(id))) +
  theme_bw() + theme(legend.position = 'none') +
  labs(x = 'Year', y = 'Counts (log-scale)')

ggplot(data = retrodata) +
  facet_wrap(~ paste0("Species ", species), scales = "free") +
  geom_point(aes(x = year, y = exp(obs), group = id, color = as.character(id)),
             size = 1) + 
  geom_line(aes(x = year, y = exp(pred), group = id, color = as.character(id))) +
  theme_bw() + theme(legend.position = 'none') +
   labs(x = 'Year', y = 'Counts')

@@

The model does a fair job at recovering population-level parameters.

```{r inference sim data}
#| eval: false
#| echo: false
#| 
species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("species", sp), "Std. Error"]}))

@@

slopes <- data.frame(obs=mu_species_beta, est=species_slope, est.sd=species_slope_sd, species = as.character(1:nspecies))
ggplot(data = slopes) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
  geom_pointrange(aes(ymin = est-est.sd, ymax = est+est.sd, x = obs, y = est, color = species)) +
  theme_bw()

```{r inference pop sim data}

em <- data.frame(emmeans::emmeans(nestlm, ~ species/population))
pops_fit$int_fit <- em$emmean
pops_fit$int_fit.lcl <- em$lower.CL
pops_fit$int_fit.ucl <- em$upper.CL

em <- data.frame(emmeans::emtrends(nestlm,  ~ species/population, var = 'year'))
pops_fit$slp_fit <- em$year.trend
pops_fit$slp_fit.lcl <- em$lower.CL
pops_fit$slp_fit.ucl <- em$upper.CL

ggplot(data = pops_fit) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
  geom_pointrange(aes(ymin = int_fit.lcl, ymax = int_fit.ucl, x = int, y = int_fit, color = as.character(species))) +
  theme_bw() +
  labs(x = 'Simulated intercepts', y = 'Estimated intercepts', colour = 'Species')

ggplot(data = pops_fit) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
  geom_pointrange(aes(ymin = slp_fit.lcl, ymax = slp_fit.ucl, x = slp, y = slp_fit, color = as.character(species))) +
  theme_bw() +
  labs(x = 'Simulated slopes', y = 'Estimated slopes', colour = 'Species')

@@

```{r lookdata}
#| eval: true

@@

@@

lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,]
lpi_subset_rsh$year <- lpi_subset_rsh$year-1980
lpi_subset_rsh$logvalue <- log(lpi_subset_rsh$value)
lpi_subset_rsh$ID <- as>character(lpi_subset_rsh$ID)
